Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-06-25 manuscript/manuscript.Rmd
minor updates.

Controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through expeirments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

@@

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018]. Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].

>>>>

Controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

@@

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023]. It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020].
